---
title: "individual2"
author: "Heather Moore"
date: "11/4/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r}
library(lubridate)
library(ggplot2)
library(dplyr)
library(tidyverse)
library(maps)


```
Individual 2
```{r}
accident <- read.csv("https://raw.githubusercontent.com/xdaiISU/ds202materials/master/hwlabs/fars2017/accident.csv", stringsAsFactors = FALSE)

person <- read.csv("https://raw.githubusercontent.com/xdaiISU/ds202materials/master/hwlabs/fars2017/person.csv", stringsAsFactors = FALSE)
```
1.) Are there some days of the week where more accidents happen than the others (see FARS manual, use variable DAY_WEEK)?
DAY_WEEK: records the day of the week on which the crash occurred.
```{r}
accident$Day_Name <- wday(accident$DAY_WEEK, label = TRUE)
acc2 <- accident %>% group_by(Day_Name) %>% summarize(dayweek = sum(DAY_WEEK))
acc2 %>% ggplot(aes(Day_Name, dayweek), width =700) + geom_point() + geom_line(group = 1) + labs(x = 'Days of the Week', y = 'Number of Accidents', title = 'Accidents Corresponding to the Days of the Week')
```
 
Analysis: In the graphic above it can be inferred that some days have more accidents than others. It appears that as the week goes on the number of accidents tend to increase. Thursday, Friday, and Saturday, the days at the end of the week,showed the highest number of accidents. In contrast, the earlier days of the week have some of the lowest number of accidents. 
  
2.) Create a data frame containing the persons who are fatally hurt in the accidents (look up variable INJ_SEV)
```{r}
per2 <- person %>% filter(INJ_SEV == '4')
head(per2)
```

3.) Create a data frame containing the most dangerous vehicle make in each state. The number of persons fatally hit in the vehicle make is used to assess the (non-)safety of a make. Make sure to handle the missing values appropriately. (look up variable MAKE)
```{r}

dangerous <- per2 %>%
  drop_na(MAKE) %>%
  group_by(STATE, MAKE) %>%
  summarise(fatal_acc = n()) %>%
  group_by(STATE) %>%
  filter(fatal_acc == max(fatal_acc)) 

head(dangerous)
```

4.Create a map, and label each state with the most dangerous vehicle. Discuss the definition of the most dangerous vehicle, and what you find from the map. (Hint: Read the description for the STATE and COUNTY columns in the FARS manual. The state & county codes are Geographic Locator Codes (GLCs) from the General Services Administration’s (GSA) publication. Use readxl::read_xlsx to read in the GLCs.)

```{r}
MakeCode <- read.csv('makeCodes.csv')
StateCode <- read.csv('stateCodes.csv')

StateCode$State<- tolower(StateCode$State)
StateCode$Codes <- as.numeric(StateCode$Codes)
dangerous$fatal_acc <- as.numeric(dangerous$fatal_acc)
dangerous$STATE <- as.numeric(dangerous$STATE)

states <- map_data("county")

head(states %>% rename("STATE" = "region"))

All <- left_join(states, dangerous, by = c('order'='STATE'))
all2 <- left_join(All, MakeCode, by = c('MAKE' = "Codes"))

all2 %>% ggplot(aes(long,lat, label = Make)) + geom_polygon(aes(group = group)) + geom_text() 
```

Analysis: It appears that the most dangerous make in the united states is Ford. That is the make that appears the most when looking at the total amount of fatal crashes.Following closely behind is Chevrolet, the next highest amount of fatal crashes. 

