---
title: "team"
author: "Adam Ford, Heather Moore"
date: "11/4/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Import
```{r}
library(lubridate)
library(ggplot2)
library(dplyr)
library(tidyverse)
library(maps)
library(ggrepel)
accident <- read.csv("https://raw.githubusercontent.com/xdaiISU/ds202materials/master/hwlabs/fars2017/accident.csv", stringsAsFactors = FALSE)
person <- read.csv("https://raw.githubusercontent.com/xdaiISU/ds202materials/master/hwlabs/fars2017/person.csv", stringsAsFactors = FALSE)
```
## 1.
Are there some days of the week where more accidents happen than the others (see FARS manual, use variable DAY_WEEK)?
DAY_WEEK: records the day of the week on which the crash occurred.
```{r}
accident$Day_Name <- wday(accident$DAY_WEEK, label = TRUE)
acc2 <- accident %>% group_by(Day_Name) %>% summarize(dayweek = sum(DAY_WEEK))
acc2 %>% ggplot(aes(Day_Name, dayweek), width =700) + geom_point() + geom_line(group = 1) + labs(x = 'Days of the Week', y = 'Number of Accidents', title = 'Accidents Corresponding to the Days of the Week')
```
 
Analysis: In the graphic above it can be inferred that some days have more accidents than others. It appears that as the week goes on the number of accidents tend to increase. Thursday, Friday, and Saturday, the days at the end of the week,showed the highest number of accidents. In contrast, the earlier days of the week have some of the lowest number of accidents. 
  
## 2. 
Create a data frame containing the persons who are fatally hurt in the accidents (look up variable INJ_SEV)
```{r}
per2 <- person %>% filter(INJ_SEV == '4')
head(per2)
```

## 3.
Create a data frame containing the most dangerous vehicle make in each state. The number of persons fatally hit in the vehicle make is used to assess the (non-)safety of a make. Make sure to handle the missing values appropriately. (look up variable MAKE)
```{r}
dangerous <- per2 %>%
  drop_na(MAKE) %>%
  group_by(STATE, MAKE) %>%
  summarise(fatal_acc = n()) %>%
  group_by(STATE) %>%
  filter(fatal_acc == max(fatal_acc)) 


head(dangerous, 10)
```

## 4.
Create a map, and label each state with the most dangerous vehicle. Discuss the definition of the most dangerous vehicle, and what you find from the map. (Hint: Read the description for the STATE and COUNTY columns in the FARS manual. The state & county codes are Geographic Locator Codes (GLCs) from the General Services Administration’s (GSA) publication. Use readxl::read_xlsx to read in the GLCs.)

```{r}
MakeCode <- read.csv('makeCodes.csv')
StateCode <- read.csv('stateCodes.csv')

StateCode$State<- tolower(StateCode$State)
StateCode$Codes <- as.numeric(StateCode$Codes)
dangerous$fatal_acc <- as.numeric(dangerous$fatal_acc)
dangerous$STATE <- as.numeric(dangerous$STATE)

states <- map_data("state")
labelPosition <- states %>%
   group_by(region) %>%
   summarise(long= mean(long, na.rm = TRUE), lat= mean(lat, na.rm = TRUE))
labelPosition <- left_join(labelPosition, StateCode, by = c("region" = "State"))
 head(labelPosition, 10)
labelPosition <- left_join(labelPosition, dangerous, by = c("Codes" = "STATE"))

labelPosition <- left_join(labelPosition, MakeCode, by = c("MAKE" = "Codes"))

                          
states %>% ggplot() +
  geom_polygon(  aes(x=long, y=lat, group=group),
                color="black", fill="lightblue" ) + geom_text(data=labelPosition, aes(label = Make, x=long, y=lat), size=3)
```

Analysis: It appears that the most dangerous make in the united states is Ford. That is the make that appears the most when looking at the total amount of fatal crashes.Following closely behind is Chevrolet, the next highest amount of fatal crashes.


## 5. 
Join the accident and person table (work out which variable(s) to use)
```{r}
joined <- inner_join(person, accident, by="ST_CASE")
```

  The given documentation describes that ST_CASE is the shared variable for merging between accident and person.
  
  
## 6. 
Tally the number of accidents by day of the week (DAY_WEEK), hour of the day (HOUR) and gender (SEX). Visualize the results.
```{r}
withAvgSex <- joined %>%
  filter(SEX == 1 || SEX == 2) %>%
  group_by(ST_CASE) %>%
  summarise(avgSex = mean(SEX))
withAvgSex$sexesExplained <- ifelse(withAvgSex$avgSex == 1, 'Men Only', ifelse(withAvgSex$avgSex == 2, 'Women Only', 'Men and Women'))

accident <- withAvgSex %>%
  select(ST_CASE, sexesExplained) %>%
  inner_join( accident, by='ST_CASE')

grouped <- accident %>%
  filter(HOUR <= 24) %>%
  filter(DAY_WEEK != 9) %>%
  group_by(sexesExplained, HOUR, DAY_WEEK) %>%
  summarise(numAccidents = n()) %>%
  arrange(desc(numAccidents))
head(grouped, 10)
```

  From this table we begin to understand that the weekends have the most accidents, especially later at night.
  
```{r}
ggplot(grouped,aes(x=HOUR, y=numAccidents)) + geom_bar(stat='identity') + facet_grid(sexesExplained~DAY_WEEK)+xlab("Number of accidents")+ylab("Hour of day")
```

  This graph shows several factors off the count of accidents. First off, we see the number of accidents thats participants are only female is significantly smaller than the number of accidents with only men. As expected, the number of accidents involving Men and Women is the most. This graph also shows the trends of the weekends and the night having more accidents.
  
## 7. 
Now plot a choropleth map of the number of deaths on a county level.
```{r}
deathCounty <- person %>%
  filter(INJ_SEV == 4) %>%
  group_by(STATE, COUNTY) %>%
  summarise(numDeaths = n())
counties <- map_data("county")
countyStateCodes <- readr::read_csv('statesCounties.csv')
countyStateCodes$StateName <- tolower(countyStateCodes$StateName)
countyStateCodes$CountyName <- tolower(countyStateCodes$CountyName)
countyStateCodes$StateCode <- as.numeric(countyStateCodes$StateCode)
countyStateCodes$CountyCode <- as.numeric(countyStateCodes$CountyCode)
deathCounty$STATE <- as.numeric(deathCounty$STATE)
deathCounty$COUNTY <- as.numeric(deathCounty$COUNTY)
withNames <- right_join(deathCounty, countyStateCodes, by = c("STATE" = "StateCode", "COUNTY" = "CountyCode"))
withNames$numDeaths <- replace_na(withNames$numDeaths, 0)
 allTogether <- inner_join(withNames, counties, by = c("StateName" = "region", "CountyName" = "subregion"))
 allTogether %>% ggplot(aes(x=long, y=lat))+geom_polygon(aes(group=group, fill=numDeaths))+scale_fill_distiller(name="Number of Deaths per county", palette = "Set3")
```

  This map shows two main hotspots of fatal accidents, California and Florida. Other metropolitan areas register as dangerous as well, but not as bad as the mentioned hot spots.
  
## 8. 
Is summer or winter more dangerous? Does this depend on states? Explore and explain.
```{r}
summer <- c(5,6,7,8)
winter <- c(12, 1, 2, 3)
accident$sumWint <- ifelse(accident$MONTH %in% summer, 'Summer', ifelse(accident$MONTH %in% winter, 'Winter', 'Neither') )
accident %>% 
  filter(sumWint == 'Summer' | sumWint == 'Winter') %>%
  group_by(sumWint) %>%
  summarise(numAcc = n()) %>%
  arrange(desc(numAcc)) %>%
  head()
```

  This shows that in the dataset there are more accidents in the Summer overall.
  
```{r}
accident <- countyStateCodes %>%
  select(StateCode, StateName) %>%
  right_join(accident, by = c("StateCode" = "STATE"))
byState <- accident %>% 
  filter(sumWint == 'Summer' | sumWint == 'Winter') %>%
  group_by(sumWint, StateName) %>%
  summarise(numAcc = n()) %>%
  arrange(desc(numAcc))
head(byState, 10)
```

  The biggest trend this shows is how many accidents Texas has a state, most likely due to its large land mass. We also start to get an idea that Summer and Winter accident are pretty equal between states.
  
  
```{r}
options(scipen=10000)
ggplot(byState, aes(fill=sumWint, x=StateName, y=numAcc))+geom_bar(position='dodge', stat='identity')+theme(axis.text.x = element_text(angle = 90))+labs(fill = "Season")+xlab("State Name")+ylab("Number of Accidents")
```

  This graph shows it does vary state by state on if Summer or Winter is more dangerous, however it is more common that Summer has more accidents.
